The Talking Heads Experiment

### 2015-04-10 chap10/chap10.tex
cumulative update, mainly for Ch. 7-12

The Talking Heads experiments were obviously not an end point, but only the beginning. They confirmed the results that 
had been obtained with earlier theoretical models and computer simulations but did not push these models 
towards greater complexity. But subsequent work did expand the envelope of our understanding and modeling way  
beyond these boundaries, thanks to the work of many researchers who have since joined the research program. 
The expansion has happened along several dimensions: increased sophistication of the robots, a deeper theoretical 
understanding of semiotic dynamics, growth in the complexity of semantics and grammar, and new breakthroughs by 
studying the emergence and evolution of language strategies. An exhaustive survey of these exciting profound developments 
is beyond the scope of the present book. Instead, I will highlight here only some of the key language game experiments that 
used real physical robots and were direct variations or further extensions of the environments, game scripts
and strategies used in the Talking Heads experiment. This chapter focuses on experiments 
in the period before 2005, particularly work with the AIBO robots and the first attempts towards the emergence 
of grammar.\footnote{An overview of these experiments is also given in: \cite{Steels:2005}.}
The next chapters focus on experiments after 2005.

\section{Experiments with the AIBO robots}

@@

\subsection{AIBO's first words}

Meanwhile significant developments started to happen in the world of robotics that pushed the state of 
the art in fully-embodied mobile robots forward.  
Around the year 2000, a new generation of robots was becoming available, built 
by industrial companies which mastered the technology to manufacture very robust machines, and these robots could therefore
be used continuously without constant breakdown. Because of experience and contacts in the field of
robotics, our team was able to move language game experiments to a new level of sophistication. The first 
platform we could use was the Sony AIBO robot. \is{AIBO}

The AIBO is a fully autonomous 4-legged mobile robot, inspired by a small dog. It is fully autonomous with more than 
a thousand behaviours coordinated through a complex behaviour-based motivational 
system.\footnote{This robot was entirely based on a series of design principles that I had pioneered together with Rodney 
Brooks in the mid-nineties under the label 'behaviour-based AI'. See: \cite{Steels:1995}, \cite{Steels:1994}.}
The robot was pioneered by
Toshi Doi and designed by Masahiro Fujita and his team at the Sony Corporation in Tokyo at the end of 
the nineties \citep{Fujita:1998}. 
In all 150,000 AIBO's were sold to customers. The AIBO featured 4-legged locomotion, 
a camera for visual input, two microphones, a wide range of body sensors, on-board batteries and the necessary 
computing power. This robot was the most complex reliable robot available in the early 2000's. 
The AIBO was also the first robot that was explicitly designed for 
human interaction so that basic useful components (for example for face recognition) were available. Language game research
could therefore focus entirely on the linguistic and conceptual aspects of the semiotic cycle, even though 
we still had to program the physical interaction patterns and behaviours necessary for language games.

Language Game experiments started on the AIBO when the Talking Heads experiment 
was still going on, namely in 2000. Fr\'{e}d\'{e}ric Kaplan was the main developer.\footnote{A paper summarising 
the AIBO experiment was published as \cite{Steels:2001}. Kaplan wrote his Ph.D thesis on the subject 
which is published (in French) as \cite{Kaplan:2001}.}

These experiments extended the Talking Heads experiment in several directions: 
\begin{enumerate}
\item {\itshape Vision.} The pan-tilt cameras and the flat white board with geometric figures ensured that the vision system 
of the Talking Heads was relatively reliable, although light conditions and camera misalignment
could occasionally 
cause havoc. The AIBO experiments now used 3d objects in the real world that were seen from many different 
angles and in different light conditions (see \figref{f:look-ball}, right). This obviously pushed up the 
difficulty of visual perception enormously and it also made it much more difficult to acquire stable categories 
for reliably recognising the objects in the environment. 
\item {\itshape Speech.} The interaction with the robot used spoken language, which implied that the robot needed 
speech synthesis and a speech recognition system that could acquire new words. These components were available on the 
robot and could therefore be easily integrated. 
\item {\itshape Flexible scripts.} The Talking Heads experiments used a single very streamlined interaction scenario whereby
the robots played a Guessing Game with a clear turn-taking script. The AIBO experiments targeted different 
kinds of interactions with humans and therefore the robot needed not only a way to represent and execute these 
different scripts but also a much more flexible way to move from one step to another within a script and to move 
between scripts. Moreover the scripts had to be integrated intimately with the physical behaviours that were 
steering the robot itself, which could in many cases not be overridden or directly controlled. 
\end{enumerate}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap10/figs/look-ball.pdf}
\includegraphics[width=.45\textwidth]{chap10/figs/views-ball.pdf}}
\caption{\label{f:look-ball} 
Left: Fr\'{e}d\'{e}ric Kaplan interacts with the AIBO robot in an experiment for the social learning of lexicons.
Right: View of the ball through the eye of the robot. As the same object is now seen from many different viewpoints it 
becomes much more challenging to consistently recognise it.}
\end{figure}
The first AIBO experiments did not involve a population of agents but rather a single agent that was interacting directly with a human
experimenter. The agent had to learn the lexicon of the human through social 
interaction within the context of situated language games. The 2001
paper ``AIBO's first words'' by Steels and Kaplan\todo{citation?} described experiments in which an AIBO robot 
learned words about objects in the environment, such as `ball' or `smiley', as well as words for actions, such as 
`sit down' or `stand up'.

A typical example of a game is the DO game, used to learn the names of behaviours.\is{DO game} 
The robot is programmed with a repertoire of behaviours, such as 
walk, sit down, stand up, look left/right/up/down, push ball, etc. When the human utters a word, the hearing robot 
looks up the word in its own memory and performs that behaviour. If it does not know the word, it 
performs one of the behaviours for which it does not have a word yet. 
If this behaviour is the wrong one, the robot receives feedback and correction, 
thus learning (i) that the behaviour is not associated with the word and (ii) what the right word is for the behaviour 
just performed. When the right behaviour was chosen, the association between this behaviour 
and the word can be stored if the association was not yet in memory, or reinforced if it was using the 
lateral inhibition learning strategy. 
Games therefore always come with two variant scripts. There is one with a successful interaction, which then 
leads to a reinforcement of the existing lexicon, and one with a non-successful interaction, where the 
robot gets corrected and learns a new word. Here are the two variants for the DO Game: 
\begin{center}
\begin{tabular}{ l  l }
\lsptoprule
{\itshape Reinforcement script}&{\itshape Correction script} \\ \midrule
Human: Listen, Walk. & H: Listen, Walk.  \\ 
R: Walk? & R: Walk? \\
 {\itshape (walks)} & {\itshape (sits)} \\
H: Yes & H: No, this is `sit'. \\
& R: Sit ? \\ 
& H: Yes. \\ 
\lspbottomrule
\end{tabular}
\end{center}
The scripts of the agents are represented internally using probabilistic finite state machines with 
temporal annotations. This computational formalism (pioneered by Rodney Brooks) was also used to program the 
physical behaviour of the robots (\figref{fig:fsm}).\footnote{
This architecture was first described in \cite{Brooks:1986}. 
For the experiments I implemented a LISP-based Behaviour Language that ran also on the AIBO.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap10/figs/fsm.pdf}}
\caption{\label{fig:fsm} 
The behaviour on the AIBO, including the language game scripts, are programmed using probabilistic finite state 
machines that bring the robot from one internal state to another one depending on internal or external conditions.}
\end{figure}

@@

The AIBO experiments were the first robotic experiments demonstrating social, situated learning in action. 
They were highly relevant for discovering the subtle interaction patterns that would work and 
could be realistically implemented in human-robot scenarios. The goal of these experiments was 
not to address the question of the origins of a 
new language in a population of agents. However, another experiment, still with the AIBO robots,  
did substantially advance the state of the art: the Perspective Reversal Experiment carried out by Martin Loetzsch 
and Luc Steels in 2004-2005.\footnote{
Very little was published about this experiment, partly because reviewers had the greatest difficulty to grasp 
its underlying rationale or understand the technical details. The main paper is \cite{Steels:2008spatial}. 
Subsequent work by Michael Spranger significantly scaled up the experiment, as discussed in the next chapter.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap10/figs/tokyo-persp.pdf}}
\caption{\label{fig:tokyo-persp} 
Luc Steels demonstrates live the Perspective Reversal Experiment in Tokyo, April 2005. AIBO robots track 
the movement of a ball and 
the robot acting as speaker describes to the hearer an observed movement, possibly taking into account the 
perspective of the hearer. For example, the speaker might say `the ball rolls from my left to your right'.}
\end{figure}

The Perspective Reversal Experiment worked as follows. A population of AIBO robots roam around
freely in an unconstrained in-door environment. As soon as one sees the
ball, it comes to a stop and searches for another robot nearby, which also
looks for the ball and stops when it sees it. Then the human
experimenter pushes the ball with a stick so that it rolls a short
distance (see \figref{fig:tokyo-persp}). The two robots next play a description game. One robot (acting as the `speaker') 
describes the ball-moving event to the other robot (the ``hearer''), going through the same semiotic cycle as the Talking 
Heads agents. The game is a success if the meaning conveyed by the speaker is compatible with the hearer's 
own perception of the scene.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.90\textwidth]{chap10/figs/persp-rev-perception.pdf}}
\caption{\label{fig:persp-rev-perception} 
Chain from visual image (left) to an analog world model (middle) and a discrete event description
that is the topic of the language game (right). 
The top row shows this chain for robot A and the bottom row the chain for robot B, both looking at the same ball.}
\end{figure}

@@

\begin{figure}[t]
\centerline{
\includegraphics[width=0.75\textwidth]{chap10/figs/stages.pdf}}
\caption{\label{fig:stages} 
Three strategies are compared. In Stage I (left) agents categorise the event from their 
own perspective. In Stage II (middle) they conceptualise the scene from the viewpoint of the other agent as well. In Stage III (right) 
they express that they performed a perspective reversal. Only the third strategy leads to successful performance.} 
\end{figure}

The Perspective Reversal experiment was significant in many ways. Not only did it use for the first time mobile robots in a 
sophisticated open real world set-up, but it also pioneered the use of perspective reversal that was the basis of later 
experiments in spatial language games with the QRIO robot by Michael Spranger and it was the first convincing example of 
the comparative method that now underlies all advanced language game research. What this experiment did not address
was how agents could come up with new strategies themselves and how more adapted strategies could propagate and become 
shared in the population.

@@

The earliest attempts to investigate the emergence of syntax date back to 1997, in other words even before 
the first Talking Heads experiment started. The first report 
appeared as a paper in the Artificial Intelligence journal in 1998 \citep{Steels:1998}. 
It was described by the journal editors (Dan Bobrow and 
Mike Brady) as ``obviously in an early stage, but very provocative and suggestive.'' The paper was based on my keynote 
lecture at IJCAI 1997 in Nagoya and used a precursor of the pan-tilt cameras built by Tony Belpaeme. 
(see \figref{f:plate7}).

@@

\begin{figure}[t]
\centerline{
\includegraphics[width=0.75\textwidth]{chap10/figs/associations.pdf}}
\caption{\label{fig:associations} 
Top: Two lexical associations have become active (association-259 and association-234) to cover the distinctive 
feature-set ([VISIB v-411] [VER v-431]) resulting from discrimination and perception. Bottom: 
The group of words and the meanings match with schemas forming part of a grammatical association that combines the words. }
\end{figure}
An application of these schemas in producing the utterance `(W O) (W U)' is represented in \figref{fig:associations} and shown 
here: 
\begin{verbatim}
127 ++> Speaker: head-40 Topic: 2 Context: 6 5 4 3 1 0
Categorial Perception:
  ([VISIB v-4111] [VER v-431])
  ([VER v-4311] [AREA v-4061])
Lexicon lookup: (Association-259 Association-234)
Syntactic structure:
  (syn-cat-77
    (syn-slot-50 (syn-cat-75 |(W O)|))
    (syn-slot-51 (syr-cat-76 |(W U)|)))
Semantic structure:
  (sem-cat-78
    (sem-slot-50 (sem-cat-75 (VER v-431)))
    (sem-slot-51 (sem-cat-76 (VISIB v-411))))
Meaning:
  ([VISIB v-4111] [VER v-4311)
Expression: ((W 0) (W U))
\end{verbatim}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap10/figs/peract-setup.pdf}}
\caption{\label{fig:peract-setup} 
The installation for the case grammar experiments used a more sophisticated vision system and objects that 
could be moved around. We see the two Talking Heads cameras in 
the background. They are oriented towards objects on the table or situations in the 
environment. The monitors show the camera images and the states of image processing, just as for the original 
Talking Heads experiment. 
}
\end{figure}

We used a similar set-up as the Talking Heads experiment (see \figref{fig:peract-setup}), but 
the world had to be made more complex, involving dynamical scenes in which 
different animate objects performed actions. So I decided to use 
puppets and enacted little scenes before the cameras (see \figref{fig:adam-eve} middle and bottom). 
There were two puppets, named Jack and Jill. They were actually sold at that time as puppet versions of Harry 
Potter and Hermione Granger inspired by the famous Harry Potter book series. 
The puppets appeared and disappeared within the field of view of the cameras. 
They manipulated small objects, for example pushing a green cube, picking it up, sliding the cube
towards the other puppet, putting it in a box, etc. To make the simulation more lively and easier to follow for 
a lay public, the agents engaging in a language game were also visually animated (see 
\figref{fig:adam-eve} top).\footnote{The animations were developed by Veronique Caraux in Paris.}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap10/figs/adam-eve.pdf}}
\centerline{\includegraphics[width=.70\textwidth]{chap10/figs/jack-give.pdf}}
\centerline{\includegraphics[width=.80\textwidth]{chap10/figs/push-jill.pdf}}
\caption{\label{fig:adam-eve} 
Top: Computer animation of the speaking and hearing agent used in public demonstrations 
of the case experiment. 
Middle: Example of a puppet scene, where Jill gives a block to Jack. It shows the source image (left) and a (partial) 
result of image processing (right). Bottom: Another example of a puppet scene where Jill pushes the block.}
\end{figure}

{\bfshape  Scaling up the vision system} 
The vision system of the Talking Heads only dealt with static scenes. This had to be scaled up to
deal with dynamical scenes and event recognition. It was the focal work of 
Jean-Christophe Baillie at the Sony Computer Science Laboratory in Paris, who built a vision system 
called PERACT \is{PERACT vision system} inspired by an event-recognition developed by Jeffrey 
Siskind.\footnote{The vision system is described in the Ph.D thesis of J-C Baillie. A summary is found here: 
\cite{Steels:2003}. Inspiration for the event recognition system came 
primarily from the paper by \cite{Siskind:2000}.}
The vision system was decomposed into three subsystems with information flowing both in a bottom-up
(from image to world model) and top-down way (from world model and predictions to image): the subsystems achieved 
object identification, event identification, and derivation of qualitative descriptors.

@@

The detection and tracking of units at different hierarchical levels starts in a bottom-up manner from the images captured by the camera, and goes through various processing steps: figure/ground separation based on color perception, creation of a cellular occupancy grid, spatial region growing to identify spatial regions, and construction of spatio-temporal continuities based on 
color histograms of the objects (\figref{fig:peract-images}).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap10/figs/peract-images.pdf}}
\caption{\label{fig:peract-images} 
Top: Some of the images recorded during the action of picking up a cylinder. 
Middle: Results of segmentation algorithms. 
Bottom: The spatio-temporal continuity of objects is tracked. 
}
\end{figure}

{\bfshape  Event Identification} 
The second subsystem detects and tracks events, again at different hierarchical levels. The task is similar
to that of detecting objects but the grouping is based on changes in properties of objects rather than 
on invariances. Event detection is organised in three steps: 
\begin{enumerate}
\item Detect change by computing qualitative descriptions for the movement of an object, the contact between 
two objects, the approaching of two objects, the positioning of an object with respect to another one, etc. 
\item Detect micro-events by grouping image subsequences during which the same configuration 
of qualitative descriptors holds. 
\item Detect events which are defined as sequences of micro-events recognised using probabilistic state machines. 
For example a pick-up event (as in \figref{fig:peract-images}) involves three micro-events: (1) the hand 
moves towards the object, (2) the hand touches the object, and (3) both move away together. Processes 
concerned with recovering such events use a library of event definitions which is matched against the stream of micro-events.
\end{enumerate}
An example of output from this subsystem is shown below. Each event has an index (e.g. 77), a time period (e.g. 
from 780 to 803), other events or properties that have played a role in recognising this event, a 
label (e.g. {\verb Appear}), and information of the status (like ongoing or finished).

@@

Not only the vision system, but also the language system had to be scaled up significantly to do these experiments. 
Together with Nicolas Neubauer, I built upon the formalism I had developed for the early syntax experiments, 
designing and implementing a next generation system, called COGAR, \is{COGAR (cognitive grammar architecture)} which stands for 
Cognitive Grammar Architecture.\footnote{
No official publication of the COGAR system exists. Such technical papers are virtually impossible to get 
published in journals or conferences. There was however a technical report: \cite{Steels:2001}.}
Here is a brief description of this formalism.

>>>>

The Talking Heads experiments were obviously not an end point, but only the beginning. They confirmed the results that 
had been obtained with earlier theoretical models and computer simulations but did not push these models 
towards greater complexity. But subsequent work did expand the envelope of our understanding and modeling way  
beyond these boundaries, thanks to the work of many researchers who have since joined the research program. 
The expansion has happened along several dimensions: increased sophistication of the robots, a deeper theoretical 
understanding of semiotic dynamics, growth in the complexity of semantics and grammar, and new breakthroughs by 
studying the emergence and evolution of language strategies. An exhaustive survey of these exciting profound developments 
is beyond the scope of the present book. Instead, I will highlight here only some of the key language game experiments that 
used real physical robots and were direct variations or further extensions of the environments, game scripts
and strategies used in the Talking Heads experiment. This chapter focuses on experiments 
in the period before 2005, particularly work with the \textsc{aibo} robots and the first attempts towards the emergence 
of grammar.\footnote{An overview of these experiments is also given in: \cite{Steels:2005}.}
The next chapters focus on experiments after 2005.

\section{Experiments with the \textsc{aibo} robots}

@@

\subsection{\textsc{aibo}'s first words}

Meanwhile significant developments started to happen in the world of robotics that pushed the state of 
the art in fully-embodied mobile robots forward.  
Around the year 2000, a new generation of robots was becoming available, built 
by industrial companies which mastered the technology to manufacture very robust machines, and these robots could therefore
be used continuously without constant breakdown. Because of experience and contacts in the field of
robotics, our team was able to move language game experiments to a new level of sophistication. The first 
platform we could use was the Sony \textsc{aibo} robot. \is{\textsc{aibo}}

The \textsc{aibo} is a fully autonomous 4-legged mobile robot, inspired by a small dog. It is fully autonomous with more than 
a thousand behaviours coordinated through a complex behaviour-based motivational 
system.\footnote{This robot was entirely based on a series of design principles that I had pioneered together with Rodney 
Brooks in the mid-nineties under the label 'behaviour-based AI'. See: \cite{Steels:1995}, \cite{Steels:1994}.}
The robot was pioneered by
Toshi Doi and designed by Masahiro Fujita and his team at the Sony Corporation in Tokyo at the end of 
the nineties \citep{Fujita:1998}. 
In all 150,000 \textsc{aibo}'s were sold to customers. The \textsc{aibo} featured 4-legged locomotion, 
a camera for visual input, two microphones, a wide range of body sensors, on-board batteries and the necessary 
computing power. This robot was the most complex reliable robot available in the early 2000's. 
The \textsc{aibo} was also the first robot that was explicitly designed for 
human interaction so that basic useful components (for example for face recognition) were available. Language game research
could therefore focus entirely on the linguistic and conceptual aspects of the semiotic cycle, even though 
we still had to program the physical interaction patterns and behaviours necessary for language games.

Language Game experiments started on the \textsc{aibo} when the Talking Heads experiment 
was still going on, namely in 2000. Fr\'{e}d\'{e}ric Kaplan was the main developer.\footnote{A paper summarising 
the \textsc{aibo} experiment was published as \cite{Steels:2001}. Kaplan wrote his Ph.D thesis on the subject 
which is published (in French) as \cite{Kaplan:2001}.}

These experiments extended the Talking Heads experiment in several directions: 
\begin{enumerate}
\item {\itshape Vision.} The pan-tilt cameras and the flat white board with geometric figures ensured that the vision system 
of the Talking Heads was relatively reliable, although light conditions and camera misalignment
could occasionally 
cause havoc. The \textsc{aibo} experiments now used 3d objects in the real world that were seen from many different 
angles and in different light conditions (see \figref{f:look-ball}, right). This obviously pushed up the 
difficulty of visual perception enormously and it also made it much more difficult to acquire stable categories 
for reliably recognising the objects in the environment. 
\item {\itshape Speech.} The interaction with the robot used spoken language, which implied that the robot needed 
speech synthesis and a speech recognition system that could acquire new words. These components were available on the 
robot and could therefore be easily integrated. 
\item {\itshape Flexible scripts.} The Talking Heads experiments used a single very streamlined interaction scenario whereby
the robots played a Guessing Game with a clear turn-taking script. The \textsc{aibo} experiments targeted different 
kinds of interactions with humans and therefore the robot needed not only a way to represent and execute these 
different scripts but also a much more flexible way to move from one step to another within a script and to move 
between scripts. Moreover the scripts had to be integrated intimately with the physical behaviours that were 
steering the robot itself, which could in many cases not be overridden or directly controlled. 
\end{enumerate}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.50\textwidth]{chap10/figs/look-ball.pdf}
\includegraphics[width=.45\textwidth]{chap10/figs/views-ball.pdf}}
\caption{\label{f:look-ball}Left: Fr\'{e}d\'{e}ric Kaplan interacts with the \textsc{aibo} robot in an experiment for the social learning of lexicons.
Right: View of the ball through the eye of the robot. As the same object is now seen from many different viewpoints it 
becomes much more challenging to consistently recognise it.}
\end{figure}
The first \textsc{aibo} experiments did not involve a population of agents but rather a single agent that was interacting directly with a human
experimenter. The agent had to learn the lexicon of the human through social 
interaction within the context of situated language games. The 2001
paper ``\textsc{aibo}'s first words'' by Steels and Kaplan \todo{citation?} described experiments in which an \textsc{aibo} robot 
learned words about objects in the environment, such as `ball' or `smiley', as well as words for actions, such as 
`sit down' or `stand up'.

A typical example of a game is the DO game, used to learn the names of behaviours.\is{DO game} 
The robot is programmed with a repertoire of behaviours, such as 
walk, sit down, stand up, look left/right/up/down, push ball, etc. When the human utters a word, the hearing robot 
looks up the word in its own memory and performs that behaviour. If it does not know the word, it 
performs one of the behaviours for which it does not have a word yet. 
If this behaviour is the wrong one, the robot receives feedback and correction, 
thus learning (i) that the behaviour is not associated with the word and (ii) what the right word is for the behaviour 
just performed. When the right behaviour was chosen, the association between this behaviour 
and the word can be stored if the association was not yet in memory, or reinforced if it was using the 
lateral inhibition learning strategy. 
Games therefore always come with two variant scripts. There is one with a successful interaction, which then 
leads to a reinforcement of the existing lexicon, and one with a non-successful interaction, where the 
robot gets corrected and learns a new word. Here are the two variants for the DO Game: 
\begin{center}
\begin{tabular}{ l  l }
\lsptoprule
{\itshape Reinforcement script}&{\itshape Correction script} \\ \midrule
Human: Listen, Walk. & H: Listen, Walk.  \\ 
R: Walk? & R: Walk? \\
 {\itshape (walks)} & {\itshape (sits)} \\
H: Yes & H: No, this is `sit'. \\
& R: Sit ? \\ 
& H: Yes. \\ 
\lspbottomrule
\end{tabular}
\end{center}
The scripts of the agents are represented internally using probabilistic finite state machines with 
temporal annotations. This computational formalism (pioneered by Rodney Brooks) was also used to program the 
physical behaviour of the robots (\figref{fig:fsm}).\footnote{
This architecture was first described in \cite{Brooks:1986}. 
For the experiments I implemented a LISP-based Behaviour Language that ran also on the \textsc{aibo}.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.70\textwidth]{chap10/figs/fsm.pdf}}
\caption{\label{fig:fsm}The behaviour on the \textsc{aibo}, including the language game scripts, are programmed using probabilistic finite state 
machines that bring the robot from one internal state to another one depending on internal or external conditions.}
\end{figure}

@@

The \textsc{aibo} experiments were the first robotic experiments demonstrating social, situated learning in action. 
They were highly relevant for discovering the subtle interaction patterns that would work and 
could be realistically implemented in human-robot scenarios. The goal of these experiments was 
not to address the question of the origins of a 
new language in a population of agents. However, another experiment, still with the \textsc{aibo} robots,  
did substantially advance the state of the art: the Perspective Reversal Experiment carried out by Martin Loetzsch 
and Luc Steels in 2004-2005.\footnote{
Very little was published about this experiment, partly because reviewers had the greatest difficulty to grasp 
its underlying rationale or understand the technical details. The main paper is \cite{Steels:2008spatial}. 
Subsequent work by Michael Spranger significantly scaled up the experiment, as discussed in the next chapter.}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.65\textwidth]{chap10/figs/tokyo-persp.pdf}}
\caption{\label{fig:tokyo-persp}Luc Steels demonstrates live the Perspective Reversal Experiment in Tokyo, April 2005. \textsc{aibo} robots track 
the movement of a ball and 
the robot acting as speaker describes to the hearer an observed movement, possibly taking into account the 
perspective of the hearer. For example, the speaker might say `the ball rolls from my left to your right'.}
\end{figure}

The Perspective Reversal Experiment worked as follows. A population of \textsc{aibo} robots roam around
freely in an unconstrained in-door environment. As soon as one sees the
ball, it comes to a stop and searches for another robot nearby, which also
looks for the ball and stops when it sees it. Then the human
experimenter pushes the ball with a stick so that it rolls a short
distance (see \figref{fig:tokyo-persp}). The two robots next play a description game. One robot (acting as the `speaker') 
describes the ball-moving event to the other robot (the ``hearer''), going through the same semiotic cycle as the Talking 
Heads agents. The game is a success if the meaning conveyed by the speaker is compatible with the hearer's 
own perception of the scene.

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.90\textwidth]{chap10/figs/persp-rev-perception.pdf}}
\caption{\label{fig:persp-rev-perception}Chain from visual image (left) to an analog world model (middle) and a discrete event description
that is the topic of the language game (right). 
The top row shows this chain for robot A and the bottom row the chain for robot B, both looking at the same ball.}
\end{figure}

@@

\begin{figure}[t]
\centerline{
\includegraphics[width=0.75\textwidth]{chap10/figs/stages.pdf}}
\caption{\label{fig:stages}Three strategies are compared. In Stage I (left) agents categorise the event from their 
own perspective. In Stage II (middle) they conceptualise the scene from the viewpoint of the other agent as well. In Stage III (right) 
they express that they performed a perspective reversal. Only the third strategy leads to successful performance.} 
\end{figure}

The Perspective Reversal experiment was significant in many ways. Not only did it use for the first time mobile robots in a 
sophisticated open real world set-up, but it also pioneered the use of perspective reversal that was the basis of later 
experiments in spatial language games with the \textsc{qrio} robot by Michael Spranger and it was the first convincing example of 
the comparative method that now underlies all advanced language game research. What this experiment did not address
was how agents could come up with new strategies themselves and how more adapted strategies could propagate and become 
shared in the population.

@@

The earliest attempts to investigate the emergence of syntax date back to 1997, in other words even before 
the first Talking Heads experiment started. The first report 
appeared as a paper in the Artificial Intelligence journal in 1998 \citep{Steels:1998}. 
It was described by the journal editors (Dan Bobrow and 
Mike Brady) as ``obviously in an early stage, but very provocative and suggestive.'' The paper was based on my keynote 
lecture at \textsc{ijcai} 1997 in Nagoya and used a precursor of the pan-tilt cameras built by Tony Belpaeme. 
(see \figref{f:plate7}).

@@

\begin{figure}[t]
\centerline{
\includegraphics[width=0.75\textwidth]{chap10/figs/associations.pdf}}
\caption{\label{fig:associations}Top: Two lexical associations have become active (association-259 and association-234) to cover the distinctive 
feature-set ([VISIB v-411] [VER v-431]) resulting from discrimination and perception. Bottom: 
The group of words and the meanings match with schemas forming part of a grammatical association that combines the words. }
\end{figure}
An application of these schemas in producing the utterance `(W O) (W U)' is represented in \figref{fig:associations} and shown 
here: 
\begin{verbatim}
127 ++> Speaker: head-40 Topic: 2 Context: 6 5 4 3 1 0
Categorial Perception:
  ([VISIB v-4111] [VER v-431])
  ([VER v-4311] [AREA v-4061])
Lexicon lookup: (Association-259 Association-234)
Syntactic structure:
  (syn-cat-77
    (syn-slot-50 (syn-cat-75 |(W O)|))
    (syn-slot-51 (syr-cat-76 |(W U)|)))
Semantic structure:
  (sem-cat-78
    (sem-slot-50 (sem-cat-75 (VER v-431)))
    (sem-slot-51 (sem-cat-76 (VISIB v-411))))
Meaning:
  ([VISIB v-4111] [VER v-4311)
Expression: ((W 0) (W U))
\end{verbatim}

@@

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.60\textwidth]{chap10/figs/PERACT-setup.pdf}}
\caption{\label{fig:PERACT-setup}The installation for the case grammar experiments used a more sophisticated vision system and objects that 
could be moved around. We see the two Talking Heads cameras in 
the background. They are oriented towards objects on the table or situations in the 
environment. The monitors show the camera images and the states of image processing, just as for the original 
Talking Heads experiment. 
}
\end{figure}

We used a similar set-up as the Talking Heads experiment (see \figref{fig:PERACT-setup}), but 
the world had to be made more complex, involving dynamical scenes in which 
different animate objects performed actions. So I decided to use 
puppets and enacted little scenes before the cameras (see \figref{fig:adam-eve} middle and bottom). 
There were two puppets, named Jack and Jill. They were actually sold at that time as puppet versions of Harry 
Potter and Hermione Granger inspired by the famous Harry Potter book series. 
The puppets appeared and disappeared within the field of view of the cameras. 
They manipulated small objects, for example pushing a green cube, picking it up, sliding the cube
towards the other puppet, putting it in a box, etc. To make the simulation more lively and easier to follow for 
a lay public, the agents engaging in a language game were also visually animated (see 
\figref{fig:adam-eve} top).\footnote{The animations were developed by Veronique Caraux in Paris.}

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.40\textwidth]{chap10/figs/adam-eve.pdf}}
\centerline{\includegraphics[width=.70\textwidth]{chap10/figs/jack-give.pdf}}
\centerline{\includegraphics[width=.80\textwidth]{chap10/figs/push-jill.pdf}}
\caption{\label{fig:adam-eve}Top: Computer animation of the speaking and hearing agent used in public demonstrations 
of the case experiment. 
Middle: Example of a puppet scene, where Jill gives a block to Jack. It shows the source image (left) and a (partial) 
result of image processing (right). Bottom: Another example of a puppet scene where Jill pushes the block.}
\end{figure}

{\bfshape  Scaling up the vision system} 
The vision system of the Talking Heads only dealt with static scenes. This had to be scaled up to
deal with dynamical scenes and event recognition. It was the focal work of 
Jean-Christophe Baillie at the Sony Computer Science Laboratory in Paris, who built a vision system 
called \textsc{peract} \is{\textsc{peract} vision system} inspired by an event-recognition developed by Jeffrey 
Siskind.\footnote{The vision system is described in the Ph.D thesis of J-C Baillie. A summary is found here: 
\cite{Steels:2003}. Inspiration for the event recognition system came 
primarily from the paper by \cite{Siskind:2000}.}
The vision system was decomposed into three subsystems with information flowing both in a bottom-up
(from image to world model) and top-down way (from world model and predictions to image): the subsystems achieved 
object identification, event identification, and derivation of qualitative descriptors.

@@

The detection and tracking of units at different hierarchical levels starts in a bottom-up manner from the images captured by the camera, and goes through various processing steps: figure/ground separation based on color perception, creation of a cellular occupancy grid, spatial region growing to identify spatial regions, and construction of spatio-temporal continuities based on 
color histograms of the objects (\figref{fig:PERACT-images}).

\begin{figure}[htbp]
  \centerline{\includegraphics[width=.80\textwidth]{chap10/figs/PERACT-images.pdf}}
\caption{\label{fig:PERACT-images}Top: Some of the images recorded during the action of picking up a cylinder. 
Middle: Results of segmentation algorithms. 
Bottom: The spatio-temporal continuity of objects is tracked. 
}
\end{figure}

{\bfshape  Event Identification} 
The second subsystem detects and tracks events, again at different hierarchical levels. The task is similar
to that of detecting objects but the grouping is based on changes in properties of objects rather than 
on invariances. Event detection is organised in three steps: 
\begin{enumerate}
\item Detect change by computing qualitative descriptions for the movement of an object, the contact between 
two objects, the approaching of two objects, the positioning of an object with respect to another one, etc. 
\item Detect micro-events by grouping image subsequences during which the same configuration 
of qualitative descriptors holds. 
\item Detect events which are defined as sequences of micro-events recognised using probabilistic state machines. 
For example a pick-up event (as in \figref{fig:PERACT-images}) involves three micro-events: (1) the hand 
moves towards the object, (2) the hand touches the object, and (3) both move away together. Processes 
concerned with recovering such events use a library of event definitions which is matched against the stream of micro-events.
\end{enumerate}
An example of output from this subsystem is shown below. Each event has an index (e.g. 77), a time period (e.g. 
from 780 to 803), other events or properties that have played a role in recognising this event, a 
label (e.g. {\verb Appear}), and information of the status (like ongoing or finished).

@@

Not only the vision system, but also the language system had to be scaled up significantly to do these experiments. 
Together with Nicolas Neubauer, I built upon the formalism I had developed for the early syntax experiments, 
designing and implementing a next generation system, called \textsc{cogar}, \is{\textsc{cogar} (cognitive grammar architecture)} which stands for 
Cognitive Grammar Architecture.\footnote{
No official publication of the \textsc{cogar} system exists. Such technical papers are virtually impossible to get 
published in journals or conferences. There was however a technical report: \cite{Steels:2001}.}
Here is a brief description of this formalism.
